paper figures

### 2025-01-09 paper_figures.Rmd
add Vincendeau & HTGTS; re-number figures

---
title: "paper figures"
author: "Benedikt Obermayer"
date: "2024/12/10"
output: 
  html_document:
    toc: true
    df_print: paged
    code_folding: hide
---

@@

genotype_colors <- c('WT'='gray40','Xlf'='#e1c2f0','Shld1'='#b870dc','Shld1_Xlf'='#a447d2',
                     'Trp53bp1 KO'='darksalmon','Brca1 KO'='darkkhaki','Rif1 KO'='skyblue1','Lig4 KO'='mediumpurple2')

@@

<!-- # get data -->

<!-- ```{r get_data} -->
<!-- # data from Vincendeau for Fig. 2 -->
<!-- # mouse tissue data for Figs. 2 + 3  -->
<!-- # CH12 data for Fig. 3 -->
<!-- # patient data for Fig. 4 -->
<!-- FB.data <- read.csv('data/FB_data.csv',header=1, row.names=1) -->
<!-- FB.data.pooled <- read.csv('data/FB_data_pooled.csv',header=1, row.names=1) -->
<!-- QP.data <- read.csv('data/QP_data.csv',header=1, row.names=1) -->
<!-- ``` -->

@@

```{r FigS3A,fig.width=3.5 * fig_scale,fig.height=1.7 * fig_scale}
HD.var %>%  
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(variable, levels=c('donor','sex','age','batch','sequencing','Residuals'))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('library','isotypes','diversity','structural','context','breaks','variants')) %>%
  ggplot(aes(x=category, y=100*variance, fill=variable)) + 
  geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA, lwd=.5) + 
  geom_jitter(position=position_jitterdodge(jitter.width=.3,dodge.width=.8), size=.25) +
  my_theme() +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        strip.clip='off',
        legend.position='top',
        legend.box='horizontal') +
  scale_fill_manual(values=var_colors) +
  labs(x='', y='% explained variance', fill='')
```

@@

```{r FigS3D_1, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
tmp <- HD.var %>%  
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(gsub('log10_nreads','# reads',variable),
                                levels=rev(c('donor','batch','sequencing','# reads','Residuals')))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('diversity','diversity_raw'))

@@

```{r FigS3D_2, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
tmp <- HD.var %>%  
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(gsub('log10_nreads','# reads',variable),
                                levels=rev(c('donor','cohort','batch','sequencing','# reads','Residuals')))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('context','context_isotype'), variable!='cohort')

@@

```{r FigS3D_3, fig.width=1.4 * fig_scale,fig.height=1.7 * fig_scale}
HD.data.br <- read.csv('data/HD_data_by_reads.csv', header=1, row.names=1)

@@

```{r FigS3B, fig.width=4.25 * fig_scale, fig.height=4.25 * fig_scale}
library(dendextend)
FB.X <- HD.data %>%
  dplyr::filter(cohort=='C2') %>%
  dplyr::select(sample, feature, value) %>%
  spread(feature, value) %>%
  dplyr::mutate_if(is.numeric, scale) %>%
  tibble::column_to_rownames('sample')

@@

```{r FigS3E, fig.width=1.5 * fig_scale, fig.height=1.75 * fig_scale}
cols.here.mouse <- c('mean_length_SG')
cohort.data <- rbind(mouse.data %>%
                       dplyr::mutate(species='mouse tissue',
                                     cohort=material) %>%
                       dplyr::select(species, cohort, cols.here.mouse),
                     vincendeau.data %>%
                       dplyr::mutate(species='Vincendeau',
                                     cohort=genotype) %>%
                       dplyr::select(species, cohort, cols.here.mouse)) %>%
  dplyr::mutate(cohort=factor(cohort, levels=c('lymph node','splenocytes','WT','Xlf','Shld1','Shld1_Xlf'))) %>%
  gather(variable, value, cols.here.mouse) %>%
  dplyr::select(variable, value, species, cohort) %>%
  dplyr::left_join(columns, by=c('variable'='col')) %>%
  dplyr::mutate(variable=factor(gsub('_downsampled|_switch','',variable), 
                                levels=gsub('_downsampled|_switch','',cols.here.mouse))) %>%
  dplyr::filter(category %in% c('diversity','isotypes','breaks','context','structural'))

@@

```{r FigS3C, fig.width=2.75 * fig_scale, fig.height=5 * fig_scale}
tmp <- HD.scaled %>%
  dplyr::filter(feature %in% selected.cols, batch=='20230313') %>%
  dplyr::select(feature,value,sequencing,donor) %>%
  spread(sequencing,value) %>% 
  dplyr::left_join(columns,by=c('feature'='col')) %>% 
  dplyr::group_by(feature) %>%
  dplyr::mutate(R=cor(minION,pacBio)) %>%
  dplyr::mutate(feature=factor(feature, levels=columns %>% 
                                 dplyr::filter(col %in% selected.cols) %>% 
                                 dplyr::pull(col)),
                feature_name=gsub('_downsampled|_switch','',feature)) %>% #,' (R=',signif(R,2),')')) %>%
  dplyr::arrange(feature)

cols_shapes <- tmp %>% 
  dplyr::distinct(category, feature, feature_name) %>%
  dplyr::mutate(var_col=cat_cols[as.character(category)]) %>%
  dplyr::group_by(category) %>%
  dplyr::mutate(var_ind=1:n())

ggplot(tmp,aes(x=minION,y=pacBio)) +
  geom_abline(slope=1,intercept=0,color='gray',linewidth=.5) +
  geom_point(aes(color=feature,shape=feature),size=1) + 
  my_theme() +
  stat_cor(aes(label = ..r.label..), size=2.5) +
  labs(color='',shape='') +
  scale_color_manual(values=cols_shapes %>%
                       dplyr::pull(var_col,feature),
                     labels=cols_shapes %>%
                       dplyr::pull(feature_name)) +
  scale_shape_manual(values=cols_shapes %>%
                       dplyr::pull(var_ind,feature),
                     labels=cols_shapes %>%
                       dplyr::pull(feature_name)) +
  guides(color=guide_legend(ncol=2),
         shape=guide_legend(ncol=2))  +
  theme(legend.position='bottom',
        legend.key.size=unit(1.5,'mm'),
        legend.key.spacing.y=unit(0, 'mm'))
```

@@

cols_shapes <- tmp %>% 
  dplyr::distinct(category, variable) %>%
  dplyr::mutate(var_col=cat_cols[as.character(category)],
                variable_name=gsub('_downsampled','',variable)) %>%
  dplyr::group_by(category) %>%
  dplyr::mutate(var_ind=1:n())

@@

@@

@@

ggplot(pcadat,aes(x=PC1,y=PC2, color=genotype)) +
  geom_point(size=2) +
  geom_polygon(data = pcadat %>%
                 dplyr::group_by(genotype) %>%
                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=genotype)) +
  my_theme() +
  theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
        legend.position='none') +
  labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
       y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
       color='',shape='',fill='') +
  scale_fill_manual(values=genotype_colors) +
  scale_color_manual(values=genotype_colors)

# ggplot(pcadat,aes(x=PC3,y=PC4, color=genotype)) +
#   geom_point(size=2) +
#   geom_polygon(data = pcadat %>%
#                  dplyr::group_by(genotype) %>%
#                  dplyr::slice(chull(PC3, PC4)), alpha = 0.25, aes(fill=genotype)) +
#   my_theme() +
#   theme(aspect.ratio=pca$sdev[4]**2/pca$sdev[3]**2) +
#   labs(x=paste0('PC 3 (',100*signif(pca$sdev[3]**2/sum(pca$sdev**2),3),'%)'),
#        y=paste0('PC 4 (',100*signif(pca$sdev[4]**2/sum(pca$sdev**2),3),'%)'),
#        color='',shape='',fill='') +
#   scale_fill_manual(values=genotype_colors) +
#   scale_color_manual(values=genotype_colors)
# 
# ggplot(pcadat,aes(x=PC1,y=PC2, color=batch)) +
#   geom_point(size=2) +
#   my_theme() +
#   theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2) +
#   labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
#        y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
#        color='',shape='')

@@

```{r FigS3G, fig.width=1. * fig_scale, fig.height=1.75 * fig_scale}
batch.CV <- c()
for (batch in unique(CH12.meta$batch)) {
  train <- CH12.meta$batch != batch
  test <- CH12.meta$batch == batch
  X <- CH12.data[, CH12.cols] %>%
    dplyr::mutate_if(is.numeric, function(x) { 
      s <- numeric(sum(train | test)); 
      s[train] <- (x[train]-mean(x[train]))/sd(x[train]);
      s[test] <- (x[test]-mean(x[test]))/sd(x[test]);
      s
      }) %>%
    as.matrix()

@@

@@

ggplot(pcadat,aes(x=PC1,y=PC2)) +
  geom_point(aes(color=diagnosis),size=1) +
  geom_line(aes(group=donor, color=diagnosis)) +
  geom_polygon(data = pcadat %>%
                 dplyr::group_by(status) %>%
                 dplyr::slice(chull(PC1, PC2)), alpha = 0.1, aes(fill=status), legend=FALSE) +
  my_theme() +
  #geom_label_repel(aes(label=gsub('FB_','',donor)),size=2, segment.size=.25, label.padding = unit(.5,'mm'), force=20, show.legend=FALSE) +
  scale_fill_manual(values=diagnosis_colors, guide="none") +
  scale_color_manual(values=diagnosis_colors) +
  guides(color=guide_legend(override.aes = list(shape=15,size=2), ncol=3)) +
  coord_cartesian(xlim=c(-8,10),
                  ylim=c(-5,8),
                  clip='off') +
  theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
        legend.position='bottom',
        legend.box='horizontal') +
  labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
       y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
       color='',shape='',fill='')
```

@@

project new data into the old PCA

>>>>

---
title: "paper figures"
author: "Benedikt Obermayer"
date: "2024/12/19"
output: 
  html_document:
    toc: true
    df_print: paged
    code_folding: hide
---

@@

genotype_colors <- c('WT'='gray40','Xlf'='#e1c2f0','Shld1'='#b870dc','Shld1_Xlf'='#a447d2',
                     'Trp53bp1 KO'='darksalmon','Brca1 KO'='darkkhaki','Rif1 KO'='skyblue1','Lig4 KO'='mediumpurple2', 
                     'Trp53bp1'='darksalmon','Atm'='#A65628','Lig4'='mediumpurple2', 'Rif1'='skyblue1', 'Brca1'='darkkhaki')

@@

@@

we now have `r length(selected.cols)` robust features

```{r FigS2F,fig.width=3.5 * fig_scale,fig.height=1.7 * fig_scale}
HD.var %>%  
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(variable, levels=c('donor','sex','age','batch','sequencing','Residuals'))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('library','isotypes','diversity','structural','context','breaks','variants')) %>%
  ggplot(aes(x=category, y=100*variance, fill=variable)) + 
  geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA, lwd=.5) + 
  geom_jitter(position=position_jitterdodge(jitter.width=.3,dodge.width=.8), size=.25) +
  my_theme() +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        strip.clip='off',
        legend.position='top',
        legend.box='horizontal') +
  scale_fill_manual(values=var_colors) +
  labs(x='', y='% explained variance', fill='')
```

@@

```{r FigS2J_1, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
tmp <- HD.var %>%  
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(gsub('log10_nreads','# reads',variable),
                                levels=rev(c('donor','batch','sequencing','# reads','Residuals')))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('diversity','diversity_raw'))

@@

```{r FigS2J_2, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
tmp <- HD.var %>%  
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(gsub('log10_nreads','# reads',variable),
                                levels=rev(c('donor','cohort','batch','sequencing','# reads','Residuals')))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('context','context_isotype'), variable!='cohort')

@@

```{r FigS2J_3, fig.width=1.4 * fig_scale,fig.height=1.7 * fig_scale}
HD.data.br <- read.csv('data/HD_data_by_reads.csv', header=1, row.names=1)

@@

```{r FigS2H, fig.width=4.25 * fig_scale, fig.height=4.25 * fig_scale}
library(dendextend)
FB.X <- HD.data %>%
  dplyr::filter(cohort=='C2') %>%
  dplyr::select(sample, feature, value) %>%
  spread(feature, value) %>%
  dplyr::mutate_if(is.numeric, scale) %>%
  tibble::column_to_rownames('sample')

@@

```{r FigS2K, fig.width=1.5 * fig_scale, fig.height=1.75 * fig_scale}
cols.here.mouse <- c('mean_length_SG')
cohort.data <- rbind(mouse.data %>%
                       dplyr::mutate(species='mouse tissue',
                                     cohort=material) %>%
                       dplyr::select(species, cohort, cols.here.mouse),
                     vincendeau.data %>%
                       dplyr::mutate(species='Vincendeau',
                                     cohort=genotype) %>%
                       dplyr::select(species, cohort, cols.here.mouse)) %>%
  dplyr::mutate(cohort=factor(cohort, levels=c('lymph node','splenocytes','WT','Xlf','Shld1','Shld1_Xlf'))) %>%
  gather(variable, value, cols.here.mouse) %>%
  dplyr::select(variable, value, species, cohort) %>%
  dplyr::left_join(columns, by=c('variable'='col')) %>%
  dplyr::mutate(variable=factor(gsub('_downsampled|_switch','',variable), 
                                levels=gsub('_downsampled|_switch','',cols.here.mouse))) %>%
  dplyr::filter(category %in% c('diversity','isotypes','breaks','context','structural'))

@@

```{r FigS2G, fig.width=2.75 * fig_scale, fig.height=5 * fig_scale}
tmp <- HD.scaled %>%
  dplyr::filter(feature %in% selected.cols, batch=='20230313') %>%
  dplyr::select(feature,value,sequencing,donor) %>%
  spread(sequencing,value) %>% 
  dplyr::left_join(columns,by=c('feature'='col')) %>% 
  dplyr::group_by(feature) %>%
  dplyr::mutate(R=cor(minION,pacBio)) %>%
  dplyr::mutate(feature=factor(feature, levels=columns %>% 
                                 dplyr::filter(col %in% selected.cols) %>% 
                                 dplyr::pull(col)),
                feature_name=gsub('_downsampled|_switch','',feature)) %>% #,' (R=',signif(R,2),')')) %>%
  dplyr::arrange(feature)

cols_shapes <- tmp %>% 
  dplyr::distinct(category, feature, feature_name) %>%
  dplyr::mutate(var_col=cat_cols[as.character(category)]) %>%
  dplyr::group_by(category) %>%
  dplyr::mutate(var_ind=1:n()) %>%
  dplyr::mutate(feature=factor(feature, levels=levels(tmp$feature))) %>% 
  dplyr::arrange(feature)

ggplot(tmp,aes(x=minION,y=pacBio)) +
  geom_abline(slope=1,intercept=0,color='gray',linewidth=.5) +
  geom_point(aes(color=feature,shape=feature),size=1) + 
  #geom_text_repel(aes(label=feature), size=1) +
  my_theme() +
  stat_cor(aes(label = ..r.label..), size=2.5) +
  labs(color='',shape='') +
  scale_color_manual(values=cols_shapes %>%
                       dplyr::pull(var_col,feature),
                     labels=cols_shapes %>%
                       dplyr::pull(feature_name)) +
  scale_shape_manual(values=cols_shapes %>%
                       dplyr::pull(var_ind,feature),
                     labels=cols_shapes %>%
                       dplyr::pull(feature_name)) +
  guides(color=guide_legend(ncol=2),
         shape=guide_legend(ncol=2))  +
  theme(legend.position='bottom',
        legend.key.size=unit(1.5,'mm'),
        legend.key.spacing.y=unit(0, 'mm'))
```

@@

cols_shapes <- tmp %>% 
  dplyr::distinct(category, variable) %>%
  dplyr::mutate(var_col=cat_cols[as.character(category)],
                variable_name=gsub('_downsampled','',variable)) %>%
  dplyr::group_by(category) %>%
  dplyr::mutate(var_ind=1:n()) %>%
  dplyr::mutate(variable=factor(variable, levels=levels(tmp$variable))) %>% 
  dplyr::arrange(variable)

@@

```{r FigS3A, fig.width=5.5 * fig_scale, fig.height=4.5 * fig_scale}
HTGTS.data <- read.csv("data/HTGTS_data.csv", header=1, row.names=1) %>%
  tibble::column_to_rownames('sample')

HTGTS.cols <- intersect(colnames(HTGTS.data),selected.cols.mouse)

tmp <- HTGTS.data %>%
  gather(variable, value, HTGTS.cols) %>%
  dplyr::group_by(variable, celltype) %>%
  dplyr::mutate(value=scale(value)) %>%
  dplyr::group_by(celltype, genotype, variable) %>%
  dplyr::summarise(mean=mean(value),
                   se=sd(value)/sqrt(n())) %>%
  tidyr::pivot_wider(id_cols=c('variable'),names_from=c('celltype','genotype'),values_from=c('mean','se')) %>%
  dplyr::left_join(columns,by=c('variable'='col')) %>%
  dplyr::group_by(variable) %>%
  dplyr::mutate(variable=factor(variable, levels=columns %>% 
                                  dplyr::filter(col %in% HTGTS.cols) %>% 
                                  dplyr::pull(col)),
                var_col=cat_cols[category]) %>%
  dplyr::arrange(variable)

cols_shapes <- tmp %>% 
  dplyr::distinct(category, variable) %>%
  dplyr::mutate(var_col=cat_cols[as.character(category)],
                variable_name=gsub('_downsampled','',variable)) %>%
  dplyr::group_by(category) %>%
  dplyr::mutate(var_ind=1:n()) %>%
  dplyr::mutate(variable=factor(variable, levels=levels(tmp$variable))) %>% 
  dplyr::arrange(variable)

comparisons <- list(c('CH12','WT','Atm'),
                    c('CH12','WT','Trp53bp1'),
                    c('CH12','WT','Lig4'),
                    c('spleen','WT','Atm'),
                    c('spleen','WT','Trp53bp1'))

plots <- lapply(comparisons, function(x) 
  tmp %>%
    dplyr::mutate_(mean1=paste0('mean_',x[[1]],'_',x[[2]]),
                   mean2=paste0('mean_',x[[1]],'_',x[[3]]),
                   se1=paste0('se_',x[[1]],'_',x[[2]]),
                   se2=paste0('se_',x[[1]],'_',x[[3]])) %>%
    dplyr::mutate(tstat=abs(mean2-mean1)/(sqrt(se2**2+se1**2)),
                  pval=dt(tstat,4)) %>%
    ggplot(aes(x=mean1,y=mean2)) +
    geom_abline(slope=1,intercept=0,color='gray',linewidth=.5) +
    geom_point(aes(color=variable,shape=variable,size=ifelse(pval < .05,'sig','ns'))) + 
    stat_cor(method = "pearson", aes(label=..r.label..), size=2.5, show.legend=FALSE, label.y.npc = 'bottom') +
    geom_errorbar(aes(ymin=mean2-se2,
                      ymax=mean2+se2,
                      color=variable),
                  size=.25) +
    geom_errorbarh(aes(xmin=mean1-se1,
                       xmax=mean1+se1,
                       color=variable),
                   size=.25) +
    my_theme() +
    labs(color='',shape='') +
    scale_color_manual(values=cols_shapes %>% 
                         dplyr::pull(var_col,variable),
                       labels=cols_shapes %>%
                         dplyr::pull(variable_name)) +
    scale_shape_manual(values=cols_shapes %>%
                         dplyr::pull(var_ind,variable),
                       labels=cols_shapes %>%
                         dplyr::pull(variable_name)) +
    scale_size_manual(values=c('sig'=2,'ns'=.5),guide='none') +
    guides(color=guide_legend(ncol=4),
           shape=guide_legend(ncol=4)) +
    labs(x=x[[2]],y=x[[3]],title=x[[1]],fill='',color='',size=''))

plot_grid(plot_grid(plotlist=lapply(plots, function(x) x + theme(legend.position='none')), 
                    ncol=3, align='vh'),
          get_legend(plots[[1]] + theme(legend.margin=margin(0,0,0,0,'mm'))), ncol=1, rel_heights=c(.65,.35))
```

```{r FigS3C, fig.width=3.75 * fig_scale, fig.height=2.5 * fig_scale}
library(gtools)

cols.here.HTGTS <- c('n_homology_switch',
                     'frac_breaks_inversions',
                     'cluster_entropy',
                     'mean_cluster_size',
                     'acceptor_score_W',
                     'donor_score_W')

HTGTS.data %>%
  dplyr::mutate(celltype=factor(celltype, levels=c('CH12','spleen')),
                genotype=factor(genotype, levels=c('WT','Atm','Lig4','Trp53bp1'))) %>%
  gather(variable, value, cols.here.HTGTS) %>%
  dplyr::select(variable, value, genotype, celltype) %>%
  dplyr::left_join(columns, by=c('variable'='col')) %>%
  dplyr::mutate(variable=factor(gsub('_downsampled|_switch','',variable), 
                                levels=gsub('_downsampled|_switch','',cols.here.HTGTS))) %>%
  dplyr::filter(category %in% c('diversity','isotypes','breaks','context','structural')) %>%
  ggplot(aes(x=genotype, y=value, fill=genotype)) + 
  geom_boxplot(position=position_dodge(width=.75), width=.7, outlier.shape=NA, lwd=.25) + 
  geom_point(position=position_jitterdodge(jitter.width=.2, dodge.width=.75, jitter.height = 0), size=.5) + 
  my_theme() +
  coord_cartesian(clip='off') +
  stat_compare_means(ref.group='WT',method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2) +
  ggh4x::facet_grid2(celltype~variable, switch='y', scales='free_y', independent='y', strip=ggh4x::strip_vanilla(clip='off')) +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        legend.position='none') +
  scale_fill_manual(values=genotype_colors) +
  #scale_x_discrete(breaks=c('lymph node','splenocytes','WT','Xlf','Shld1','Shld1_Xlf'),
  #                 limits=c('lymph node','splenocytes','','WT','Xlf','Shld1','Shld1_Xlf'), drop=FALSE) +
  labs(x='',y='', fill='') 
```

show homology values for HTGTS data

```{r FigS3D, fig.width=3.25 * fig_scale, fig.height=1.25 * fig_scale}
read.csv('data/HTGTS_homology.csv',header=1, row.names=1) %>%
  dplyr::mutate(genotype=factor(genotype, levels=c('WT','Atm','Lig4','Trp53bp1'))) %>%
  dplyr::filter(xor(isotype=='Sa', celltype=='spleen')) %>%
  dplyr::group_by( celltype, genotype, isotype, n_homology) %>%
  dplyr::summarise(mean_frac=mean(frac),
                   se_frac=sd(frac)/sqrt(n())) %>%
  ggplot(aes(x=n_homology, y=mean_frac, color=genotype)) +
  geom_line(aes(group=genotype)) +
  geom_errorbar(aes(ymin=mean_frac-se_frac,ymax=mean_frac+se_frac),width=0) +
  ggh4x::facet_nested(. ~ celltype  + isotype , scales='free', space='free_x', nest_line = TRUE, switch='y') +
  my_theme() +
  scale_color_manual(values=genotype_colors) +
  labs(color='', y='fraction junctions') +
  coord_cartesian(clip='off') 
```

```{r FigS3B, fig.width=2.5 * fig_scale, fig.height=3. * fig_scale}
HTGTS.CH12.X <- HTGTS.data[HTGTS.data$celltype=='CH12',HTGTS.cols] %>%
  dplyr::mutate_if(is.numeric, scale) 
HTGTS.spleen.X <- HTGTS.data[HTGTS.data$celltype=='spleen',HTGTS.cols] %>%
  dplyr::mutate_if(is.numeric, scale)

HTGTS.meta <- HTGTS.data %>%
  dplyr::mutate(genotype=factor(genotype,levels=c('WT','Atm','Lig4','Trp53bp1'))) %>%
  dplyr::select(celltype,genotype)

pca.CH12 <- prcomp(HTGTS.CH12.X)
pcadat.CH12 <- cbind(pca.CH12$x,HTGTS.meta[row.names(HTGTS.CH12.X),])
pca.spleen <- prcomp(HTGTS.spleen.X)
pcadat.spleen <- cbind(pca.spleen$x,HTGTS.meta[row.names(HTGTS.spleen.X),])

p1 <- ggplot(pcadat.CH12,aes(x=PC1,y=PC2, color=genotype)) +
  geom_point(size=1) +
  geom_polygon(data = pcadat.CH12 %>%
                 dplyr::group_by(genotype) %>%
                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=genotype)) +
  my_theme() +
  theme(aspect.ratio=pca.CH12$sdev[2]**2/pca.CH12$sdev[1]**2,
        legend.position=c(.92,.3),
        legend.background = element_blank()) +
  labs(x=paste0('PC 1 (',100*signif(pca.CH12$sdev[1]**2/sum(pca.CH12$sdev**2),3),'%)'),
       y=paste0('PC 2 (',100*signif(pca.CH12$sdev[2]**2/sum(pca.CH12$sdev**2),3),'%)'),
       color='',shape='',fill='', title='CH12')  +
  scale_color_manual(values=genotype_colors) +
  scale_fill_manual(values=genotype_colors)

p2 <- ggplot(pcadat.spleen,aes(x=PC1,y=PC2, color=genotype)) +
  geom_point(size=1) +
  geom_polygon(data = pcadat.spleen %>%
                 dplyr::group_by(genotype) %>%
                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=genotype)) +
  my_theme() +
  theme(aspect.ratio=pca.spleen$sdev[2]**2/pca.spleen$sdev[1]**2,
        legend.position='none',
        legend.background = element_blank()) +
  labs(x=paste0('PC 1 (',100*signif(pca.spleen$sdev[1]**2/sum(pca.spleen$sdev**2),3),'%)'),
       y=paste0('PC 2 (',100*signif(pca.spleen$sdev[2]**2/sum(pca.spleen$sdev**2),3),'%)'),
       color='',shape='',fill='', title='spleen')  +
  scale_color_manual(values=genotype_colors) +
  scale_fill_manual(values=genotype_colors)

plot_grid(p1,p2, ncol=1, align='vh')
```

these PCAs use `r dim(HTGTS.spleen.X)[2]` (spleen) and `r dim(HTGTS.CH12.X)[2]` (CH12) features

@@

```{r FigS3X1, fig.width=4 * fig_scale, fig.height=1.75 * fig_scale, include=FALSE, eval=FALSE}
dualPCA <- function(X, Y) {
  library(irlba)
  # use CCA-like extension of dual PCA as described here: https://xinmingtu.cn/blog/2022/CCA_dual_PCA/
  #sv <- irlba(as.matrix(X) %*% as.matrix(t(Y)), nv=n)
  sv <- svd(as.matrix(X) %*% as.matrix(t(Y)))
  x <- sv$u %*% diag(sqrt(sv$d))
  row.names(x) <- row.names(X)
  colnames(x) <- paste0('dPC',seq(1,dim(x)[2]))
  y <- sv$v %*% diag(sqrt(sv$d))
  row.names(y) <- row.names(Y)
  colnames(y) <- paste0('dPC',seq(1,dim(y)[2]))
  list(x=x, y=y, sdev=sqrt(sv$d/(dim(X)[2]-1)))
}

CH12.X <- CH12.data[CH12.take, mouse.cols] %>% 
  dplyr::mutate_if(is.numeric, scale)
mouse.X <- mouse.data[mouse.take, mouse.cols] %>% 
  dplyr::mutate_if(is.numeric, scale)

dpca <- dualPCA(CH12.X, mouse.X)
dpcadat <- cbind(rbind(dpca$x, dpca$y),
                 data.frame(group=c(CH12.data[row.names(dpca$x),'genotype'],
                                    mouse.data$material)))

ggplot(dpcadat,aes(x=dPC1,y=dPC2, color=group)) +
  geom_point(size=1) +
  geom_polygon(data = dpcadat %>%
                 dplyr::group_by(group) %>%
                 dplyr::slice(chull(dPC1, dPC2)), alpha = 0.1, aes(fill=group), linewidth=.1) +
  my_theme() +
  scale_fill_manual(values=c(genotype_colors,tissue_colors)) +
  scale_color_manual(values=c(genotype_colors,tissue_colors)) +
  theme(aspect.ratio=dpca$sdev[2]**2/dpca$sdev[1]**2,
        legend.spacing=unit(0,'mm')) +
  labs(x=paste0('dPC 1 (',100*signif(dpca$sdev[1]**2/sum(dpca$sdev**2),3),'%)'),
       y=paste0('dPC 2 (',100*signif(dpca$sdev[2]**2/sum(dpca$sdev**2),3),'%)'),
       color='',shape='',fill='')
```

```{r FigS3X2, fig.width=3 * fig_scale, fig.height=4 * fig_scale, include=FALSE, eval=FALSE}
tmp <- t(t(t(as.matrix(rbind(CH12.X, mouse.X))) %*% as.matrix(dpcadat[,c('dPC1','dPC2')])) / colSums(dpcadat[,c('dPC1','dPC2')]**2)) %>%
  as.data.frame() %>%
  tibble::rownames_to_column('col') %>%
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(col=factor(col, levels=columns %>% 
                             dplyr::filter(col %in% mouse.cols) %>%
                             dplyr::pull(col)),
                col_name=gsub("_downsampled|_switch",'',col))

cols_shapes <- tmp %>% 
  dplyr::distinct(category, col,col_name) %>%
  dplyr::mutate(var_col=cat_cols[as.character(category)]) %>%
  dplyr::group_by(category) %>%
  dplyr::mutate(var_ind=1:n()) %>%
  dplyr::mutate(col=factor(col, levels=levels(tmp$col))) %>% 
  dplyr::arrange(col)

ggplot(tmp,aes(x=dPC1,y=dPC2, color=col, shape=col)) +
  geom_point(size=1) +
  #geom_text_repel(aes(label=col), size=1) +
  my_theme() +
  scale_color_manual(values=cols_shapes %>%
                       dplyr::pull(var_col,col),
                     labels=cols_shapes %>%
                       dplyr::pull(col_name)) +
  scale_shape_manual(values=cols_shapes %>%
                       dplyr::pull(var_ind,col),
                     labels=cols_shapes %>%
                       dplyr::pull(col_name)) +
  theme(legend.key.size=unit(1.5,'mm'),
        aspect.ratio=dpca$sdev[2]**2/dpca$sdev[1]**2,
        legend.spacing=unit(0,'mm'),
        legend.key.spacing.y=unit(0,'mm'),
        legend.position='bottom',
        legend.box='horizontal') +
  guides(color=guide_legend(ncol=2),
         shape=guide_legend(ncol=2)) +
  labs(x='dPC 1', y='dPC 2',
       color='',shape='',fill='',alpha='')
```

@@

ggplot(pcadat,aes(x=PC1,y=PC2, color=genotype)) +
  geom_point(size=2) +
  geom_polygon(data = pcadat %>%
                 dplyr::group_by(genotype) %>%
                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=genotype)) +
  my_theme() +
  theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
        legend.position='none') +
  labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
       y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
       color='',shape='',fill='') +
  scale_fill_manual(values=genotype_colors) +
  scale_color_manual(values=genotype_colors) 
```

@@

```{r FigS3X5, fig.width=1. * fig_scale, fig.height=1.75 * fig_scale, include=FALSE}
batch.CV <- c()
for (batch in unique(CH12.meta$batch)) {
  train <- CH12.meta$batch != batch
  test <- CH12.meta$batch == batch
  X <- CH12.data[, CH12.cols] %>%
    dplyr::mutate_if(is.numeric, function(x) { 
      s <- numeric(sum(train | test)); 
      s[train] <- (x[train]-mean(x[train]))/sd(x[train]);
      s[test] <- (x[test]-mean(x[test]))/sd(x[test]);
      s
      }) %>%
    as.matrix()

@@

(average accuracy is `r mean(unlist(batch.CV))` for cross-validation across batches and `r mean(unlist(sample.CV))` for cross-validation across samples)

@@

ggplot(pcadat,aes(x=PC1,y=PC2)) +
  geom_point(aes(color=diagnosis),size=1) +
  geom_line(aes(group=donor, color=diagnosis)) +
  geom_polygon(data = pcadat %>%
                 dplyr::group_by(status) %>%
                 dplyr::slice(chull(PC1, PC2)), alpha = 0.1, aes(fill=status), legend=FALSE) +
  my_theme() +
  #geom_label_repel(aes(label=gsub('FB_','',donor)),size=2, segment.size=.25, label.padding = unit(.5,'mm'), force=20, show.legend=FALSE) +
  scale_fill_manual(values=diagnosis_colors, guide="none") +
  scale_color_manual(values=diagnosis_colors) +
  guides(color=guide_legend(override.aes = list(shape=15,size=2), ncol=3)) +
  coord_cartesian(xlim=c(-8,10),
                  ylim=c(-5,8),
                  clip='off') +
  theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
        legend.key.size=unit(1.5,'mm'),
        legend.key.spacing.y=unit(0,'mm'),
        legend.position='bottom',
        legend.box='horizontal') +
  labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
       y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
       color='',shape='',fill='')
```

(this PCA uses `r dim(human.X)[2]` features)

@@

project test data into the training data PCA
